---
title: "Board Game Analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Initialise the libraries and dataset

```{r results='hide', message=FALSE}
library(tidyverse)
library(reshape2)
board_games_raw <- read_csv('board_games.csv')
```

## Number of games published each year

There are `r nrow(board_games_raw)` games in the dataset, published over many years.  First assessment is to see how many games are in each year and how many reviews have been carried out


```{r}
p <- ggplot(board_games_raw, aes(x=year_published)) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Games published each year') +
  xlab('Year published') + ylab('')
p
```

```{r}
p <- ggplot(board_games_raw %>%
              group_by(year_published) %>%
              summarise(avg_reviews_per_game=mean(users_rated))) + 
  aes(x=year_published, y=avg_reviews_per_game) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Average reviews per game') +
  xlab('Year published') + ylab('')
p
```

There are a healthy number of reviews of games from all years, although later games have roughly twice as many reviews per game as earlier years.  A big outlier pre 1960 to be investigated

Next look at the average rating over time.

```{r}
p <- ggplot(board_games_raw %>%
              group_by(year_published) %>%
              summarise(mean_rating=mean(average_rating),
                        max_rating=max(average_rating),
                        min_rating=min(average_rating)) %>%
              gather('measure','value', 2:4)) + 
  aes(x=year_published, y=value, colour= measure) +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Max/Min and mean review score over time') +
  xlab('Year published') + ylab('')
p
```

Clearly the newer games are rated more highly, but looks like there is also a greater range of scores.  Maybe a boxplot would show these differences 

```{r}
p <- ggplot(board_games_raw ) + 
  aes(x=year_published, y=average_rating, group=year_published) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Distribution of all review scores over time') +
  xlab('Year published') + ylab('')
p
```
Finally, it's worth investigating if there is a relationship between the number of reviews and the average review score

```{r}

# remove a few outliers that have a very high number of reviews
user_reviews <- board_games_raw %>%
                filter(users_rated>100, users_rated<10000)

mean_review_score <- round(mean(user_reviews$average_rating),1)
max_reviews <- max(user_reviews$users_rated)

mean_user_reviews <- round(mean(user_reviews$users_rated),1)



p <- ggplot(user_reviews ) + 
  aes(y=users_rated, x=average_rating, group=year_published) +
  geom_point() +
  ggtitle('Size of review population giving rise to average ratings') +
  xlab('Average rating') + ylab('number of reviews') +
  geom_vline(xintercept = mean_review_score, linetype = "dashed", colour="blue") +
  annotate("text", label=paste0("mean rating: ", mean_review_score, " "),
           x=mean_review_score,  y=max_reviews, 
            hjust=1, colour="blue", size=3, colour="black") +
  geom_hline(yintercept = mean_user_reviews, linetype = "dashed", colour="blue") +
  annotate("text", label=paste0("mean number of reviews: ", mean_user_reviews, " "),
           x=0,  y=mean_user_reviews+200, 
            hjust=0, colour="blue", size=3, colour="black")
p
```

It would seem that popularity could be defined by a combination of the number of reviews and the average rating.  There are quite a few games who have a high avergae rating and a large number of reviews.

We could define the four quadrants as follows:

* *popular* higher average rating and higher number of reviews than mean
* *disappointing* higher number of reviews of mean but lower average rating
* *niche* higher average rating but lower number of reviews
* *unpopular* lower average rating and lower number of reviews than mean
 
This column can be added to the raw dataset

```{r}

# add a popularity categorisation column
board_games_raw  <- board_games_raw %>%
                mutate(popularity = if_else(average_rating>mean_review_score & users_rated > mean_user_reviews, 'popular',
                                    if_else(average_rating>mean_review_score & users_rated < mean_user_reviews, 'niche',
                                    if_else(average_rating<mean_review_score & users_rated > mean_user_reviews, 'disappointing',
                                            'unpopular'))))

```

which produces the following classification
```{r echo=FALSE, results='markup'}

table(board_games_raw$popularity)

```

## Popular games over time

We can filter on this new classification and re-examine the average scores over time of the most popular games


```{r}
p <- ggplot(board_games_raw %>%
              filter(popularity=='popular') ) + 
  aes(x=year_published, y=average_rating) +
  geom_jitter() +
  ggtitle('Most popular games over time') +
  xlab('Year published') + ylab('Average rating')
p
```

The most popular games are more recent but there is a reasonable number that have had appeal for decades.



## Categories of game

Each game is assigned one or more categories.  We can use this to assess






