---
title: "Board Game Analysis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Initialise the libraries and dataset

```{r results='hide', message=FALSE}
library(tidyverse)
library(reshape2)
board_games_raw <- read_csv('board_games.csv')
```

## Number of games published each year

There are `r nrow(board_games_raw)` games in the dataset, published over many years.  First assessment is to see how many games are in each year and how many reviews have been carried out


```{r}
p <- ggplot(board_games_raw, aes(x=year_published)) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Games published each year') +
  xlab('Year published') + ylab('')
p
```

```{r}
p <- ggplot(board_games_raw %>%
              group_by(year_published) %>%
              summarise(avg_reviews_per_game=mean(users_rated))) + 
  aes(x=year_published, y=avg_reviews_per_game) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Average reviews per game') +
  xlab('Year published') + ylab('')
p
```

There are a healthy number of reviews of games from all years, although later games have roughly twice as many reviews per game as earlier years.  A big outlier pre 1960 to be investigated

Next look at the average rating over time.

```{r}
p <- ggplot(board_games_raw %>%
              group_by(year_published) %>%
              summarise(mean_rating=mean(average_rating),
                        max_rating=max(average_rating),
                        min_rating=min(average_rating)) %>%
              gather('measure','value', 2:4)) + 
  aes(x=year_published, y=value, colour= measure) +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Max/Min and mean review score over time') +
  xlab('Year published') + ylab('')
p
```

Clearly the newer games are rated more highly, but looks like there is also a greater range of scores.  Maybe a boxplot would show these differences 

```{r}
p <- ggplot(board_games_raw ) + 
  aes(x=year_published, y=average_rating, group=year_published) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Distribution of all review scores over time') +
  xlab('Year published') + ylab('')
p
```
Finally, it's worth investigating if there is a relationship between the number of reviews and the average review score

```{r}

# remove a few outliers that have a very high number of reviews
user_reviews <- board_games_raw %>%
                filter(users_rated>100, users_rated<10000)

mean_review_score <- round(mean(user_reviews$average_rating),1)
max_reviews <- max(user_reviews$users_rated)

mean_user_reviews <- round(mean(user_reviews$users_rated),1)



p <- ggplot(user_reviews ) + 
  aes(y=users_rated, x=average_rating, group=year_published) +
  geom_point() +
  ggtitle('Size of review population giving rise to average ratings') +
  xlab('Average rating') + ylab('number of reviews') +
  geom_vline(xintercept = mean_review_score, linetype = "dashed", colour="blue") +
  annotate("text", label=paste0("mean rating: ", mean_review_score, " "),
           x=mean_review_score,  y=max_reviews, 
            hjust=1, colour="blue", size=3, colour="black") +
  geom_hline(yintercept = mean_user_reviews, linetype = "dashed", colour="blue") +
  annotate("text", label=paste0("mean number of reviews: ", mean_user_reviews, " "),
           x=0,  y=mean_user_reviews+200, 
            hjust=0, colour="blue", size=3, colour="black")
p
```

It would seem that popularity could be defined by a combination of the number of reviews and the average rating.  There are quite a few games who have a high avergae rating and a large number of reviews.

We could define the four quadrants as follows:

* *popular* higher average rating and higher number of reviews than mean
* *disappointing* higher number of reviews of mean but lower average rating
* *niche* higher average rating but lower number of reviews
* *unpopular* lower average rating and lower number of reviews than mean
 
This column can be added to the raw dataset

```{r}

# add a popularity categorisation column
board_games_raw  <- board_games_raw %>%
                mutate(popularity = if_else(average_rating>mean_review_score & users_rated > mean_user_reviews, 'popular',
                                    if_else(average_rating>mean_review_score & users_rated < mean_user_reviews, 'niche',
                                    if_else(average_rating<mean_review_score & users_rated > mean_user_reviews, 'disappointing',
                                            'unpopular'))),
                       popularity_score = if_else(popularity=='disappointing', 0,
                                          if_else(popularity=='unpopular', -1,
                                          if_else(popularity=='niche', 1,2))))

```

which produces the following classification
```{r echo=FALSE, results='markup'}

table(board_games_raw$popularity)

```

## Popular games over time

We can filter on this new classification and re-examine the average scores over time of the most popular games


```{r}
p <- ggplot(board_games_raw ) + 
  aes(x=year_published, y=popularity) +
  geom_jitter() +
  ggtitle('Popularity over time') +
  xlab('Year published') + ylab('Average rating')
p
```

The most popular games are more recent but there is a reasonable number that have had appeal for decades.





## Categories of game

Each game is assigned one or more categories.  We can use this to assess which categories contain popular games. We'll calculate an average score per category and order the categories from least to most popular



```{r}
# The category column has a variable number of comma separated values
# split the category column into the maximum number of values there can be
# and label each one catN.  Melt the data back into a single column called category

cat_columns <- paste0('cat', 
                      seq_len(max(sapply(strsplit(board_games_raw$category, ','), length))))

top_n_categories <- 50


# find the overall category popularity in the top categories

better_than_average_board_games <- board_games_raw %>% 
               separate(category, into=cat_columns, sep=',') %>%
               gather(key='cat_num', value='category', cat_columns, na.rm = TRUE) %>%
               select(-cat_num) %>%
               mutate(top_cat=fct_lump(category, top_n_categories )) %>%
               group_by(top_cat) %>%
               mutate(category_popularity=sum(popularity_score)) %>%
               select(-category) %>%
               ungroup() %>%
               arrange(category_popularity) %>%
               mutate(top_cat = factor(top_cat, unique(top_cat)))



p <- ggplot(better_than_average_board_games ) + 
  aes(x=top_cat, y=category_popularity) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Category Popularity') +
  xlab('') + ylab('Mean popularity')
p


```

Then do the same by publisher

```{r}
# The category column has a variable number of comma separated values
# split the category column into the maximum number of values there can be
# and label each one catN.  Melt the data back into a single column called category

pub_columns <- paste0('pub', 
                      seq_len(max(sapply(strsplit(board_games_raw$publisher, ','), length))))

top_n_publishers <- 50


# find the overall publisher popularity 

publisher_popularity <- board_games_raw %>% 
               separate(publisher, into=pub_columns, sep=',') %>%
               gather(key='pub_num', value='category', pub_columns, na.rm = TRUE) %>%
               select(-pub_num) %>%
               mutate(top_pub=fct_lump(category, top_n_publishers )) %>%
               group_by(top_pub) %>%
               mutate(publisher_popularity=mean(popularity_score)) %>%
               select(-category) %>%
               ungroup() %>%
               arrange(publisher_popularity) %>%
               mutate(top_pub = factor(top_pub, unique(top_pub)))



p <- ggplot(publisher_popularity ) + 
  aes(x=top_pub, y=publisher_popularity) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Category Popularity') +
  xlab('') + ylab('Mean popularity')
p


```


